---
title: "MLM HW 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Part 0 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
school_data <- read.csv("/Users/jenniferjackson/Desktop/NYU/MLM/classroom.csv")

# Part 1 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
require(lme4)
require(lmerTest)
lme1 <- lmer(mathkind ~ (1|schoolid), data = school_data)
print(summary(lme1))
ranova(lme1)
```
```{r}
# logLikelihood Values
# -6042.8
# -6119.3

# Part 2 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The estimated variance in level differences between schools is 364.3.
# The ICC is 364.3/(364.3+1344.5) = 0.21. Therefore, 21% of the variance is
# explained by between-school differneces.

# Part 3 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lme2 <- lmer(mathkind ~ ses + (1|schoolid/classid), data = school_data)
print(summary(lme2))
ranova(lme2)
```

Adding socioeconomic status, the model becomes $Y_{ijk}=b_0+b_1SES_{ijk}+\eta_{jk}+\zeta_k+\varepsilon_{ijk} ,\mbox{ with }\eta_j\sim N(0,\sigma_\eta^2),\zeta_j\sim N(0,\sigma_\zeta^2), \varepsilon_{ij} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$

Yes, the coefficient for SES (socioeconomic status) is significant because it
has a p-value under 5%. The variance decreases once adding the SES term. It
makes sense that the variance associated with school level would decrease
because schools generally contain students with similar SES. E.g., schools
in the "rich part of town", etc.

```{r, echo=TRUE}
# Part 4 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

lme3 <- lmer(mathkind ~ ses + sex + minority + (1|schoolid/classid), data = school_data)
print(summary(lme3))
ranova(lme3)
```

After adding the covariates sex and minority, the model becomes $Y_{ijk}=b_0+b_1SES_{ijk}+b_2sex_{ijk}+b_3minority_{ijk}+\eta_{jk}+\zeta_k+\varepsilon_{ijk} ,\mbox{ with }\eta_j\sim N(0,\sigma_\eta^2),\zeta_j\sim N(0,\sigma_\zeta^2), \varepsilon_{ij} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$

Based on an alpha level of 5%, the coefficient for sex is not statistically significant, but the coefficient for minority is.

Adding sex and minority to the model has slightly lowered the variance associated with school level. This makes sense since there are still some single-sex schools and, depending on the location, some schools are less ethnically diverse than others.

```{r}
# Part 5 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

lme4 <- lmer(mathkind ~ ses + sex + minority + housepov + (1|schoolid/classid), data = school_data)
print(summary(lme4))
ranova(lme4)
```
Adding the covariate housepov, the model becomes $Y_{ijk}=b_0+b_1SES_{ijk}+b_2sex_{ijk}+b_3minority_{ijk}+b_4housepov_{jk}+\eta_{jk}+\zeta_k+\varepsilon_{ijk} ,\mbox{ with }\eta_{jk}\sim N(0,\sigma_\eta^2),\zeta_j\sim N(0,\sigma_\zeta^2), \varepsilon_{ij} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$

At the alpha level of 5%, the coefficient for housepov is not statistically significant.

After adding housepov, the school-level variance has slightly decreased. 

Part 6
A student-level predictor's effect on school-level variance would be explained as a within-group difference, whereas a school-level predictor's effect, being on the same level as the variance in question, would be explained as a between-group difference.




