---
title: "HW4"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=TRUE}
# Part 0 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
dat <- read.csv("/Users/jenniferjackson/Desktop/NYU/MLM/classroom.csv")

# Part 1 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
require(lme4)
require(lmerTest)

lme1 <- lmer(mathgain ~ ses + minority + yearstea + (1|schoolid), data = dat)
print(summary(lme1))
lme1_results <- ranova(lme1)
lme1_results
logLik(lme1)
```
MODEL 1 :
$MATHGAIN_{ijk}=b_0+b_1MATHKIND_{ijk}+b_2SES_{ijk}+b_3MINORITY_{ijk}+b_4YEARSTEA_{jk}+\zeta_{0k}+\varepsilon_{ijk} ,\mbox{ with }\zeta_{0k}\sim N(0,\sigma_\zeta^2), \varepsilon_{ijk} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$

The loglikelihood is -5885.867 (df=6).

```{r, echo=TRUE}
# Part 1 (a) ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lme2 <- lmer(mathgain ~ ses + minority + yearstea + (0+yearstea|schoolid) + (1|schoolid), data = dat)
print(summary(lme2))
lme2_results <- ranova(lme2)
lme2_results
logLik(lme2)
```
MODEL 2 :
$MATHGAIN_{ijk}=b_0+b_1MATHKIND_{ijk}+b_2SES_{ijk}+b_3MINORITY_{ijk}+b_4YEARSTEA_{jk}+\zeta_{0k}+\zeta_{4k}YEARSTEA+\varepsilon_{ijk} ,\mbox{ with }\zeta_{0k}\sim N(0,\sigma_\zeta^2),\zeta_{4k}\sim N(0,\sigma_\zeta^2), \varepsilon_{ijk} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$

The loglikelihood is -5884.919 (df=7).

```{r, echo=TRUE}
# Part 1 (b) ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# LRT
anova(lme1, lme2, refit = F)
# the p-value is 0.1684.
```
Part 1 (c) and (d)
Because the p-value (0.1684) is larger than alpha = 5%, adding the yearstea random slope is not statistically significant.
```{r}
# Part 2 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lme3 <- lmer(mathgain ~ ses + minority + yearstea + (yearstea|schoolid), data = dat)
print(summary(lme3))
lme3_results <- ranova(lme3)
lme3_results
logLik(lme3)
```
The loglikelihood of Model 3 (lme3) is -5882.26 (df=8)

The model is the same. The only difference is the correlation between the zeta random effects.

MODEL 3 :
$MATHGAIN_{ijk}=b_0+b_1MATHKIND_{ijk}+b_2SES_{ijk}+b_3MINORITY_{ijk}+b_4YEARSTEA_{jk}+\zeta_{0k}+\zeta_{4k}YEARSTEA+\varepsilon_{ijk} ,\mbox{ with }\zeta_{0k}\sim N(0,\sigma_\zeta^2),\zeta_{4k}\sim N(0,\sigma_\zeta^2), \varepsilon_{ijk} \sim N(0,\sigma_\varepsilon^2),\mbox{ indep.}$
```{r}
# Part 2 (a) ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#LRT
anova(lme1, lme3, refit = F)
```
The p-value (0.02712) is less than alpha = 5%, so specifying correlation between the random effects makes for a statistically significantly better model than saying they are independent.
```{r, echo=TRUE}
# Part 3 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ranefs <- ranef(lme3)
zeta0k <- ranefs$schoolid[,1]
zeta4k <- ranefs$schoolid[,2]
plot(zeta0k,zeta4k)


```
Part 3 (b) and (c)
The estimate of the correlation between zeta0k and zeta4k is -0.78, which makes sense based on the plot above.
```{r, echo=TRUE}
# Part 4 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# List effects for schoolids 1, 2 and 3

# Model 4
lme4 <- lmer(mathgain ~ ses + minority + yearstea + (yearstea|schoolid), data = dat[dat$schoolid < 4,])
print(summary(lme4))

ranefs_4 <- ranef(lme4)
zeta0k_hat <- ranefs_4$schoolid[,1]
zeta4k_hat <- ranefs_4$schoolid[,2]

# zeta0k_hat for school 1 = 4.293084e-06
# zeta0k_hat for school 2 = -4.747614e-06
# zeta0k_hat for school 3 = 4.545298e-07

# zeta4k_hat for school 1 = -3.061557e-07
# zeta4k_hat for school 2 = 3.385699e-07
# zeta4k_hat for school 3 = -3.241420e-08

# Model 5 (for schoolid 4 only)
lme5 <- lmer(mathgain ~ ses + minority + yearstea + (yearstea|schoolid), data = dat[dat$schoolid < 5,])

ranefs_5 <- ranef(lme5)
zeta0k_hat5 <- ranefs_5$schoolid[4,1] # -76.75355
zeta4k_hat5 <- ranefs_5$schoolid[4,2] # 2.318757
```
Part 4 (b)
The coefficient for $\hat{b}_0$ in model 3 is 55.8991 and for $\hat{b}_4$ is 0.1007.
MODEL 6 :
$MATHGAIN_{ijk}=\hat{b}_0+(\hat{b}_4+\hat{\zeta}_{4k})YEARSTEA_{jk}+\zeta_{0k},\mbox{ with }\zeta_{0k}\sim N(0,\sigma_\zeta^2),\zeta_{4k}\sim N(0,\sigma_\zeta^2),\mbox{ indep.}$
